---
title: "Khaykin-Proposal.Rmd"
author: "Alex Khaykin"
date: "2022-11-10"
output: pdf_document
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# load data
pres_elect <- read_rds("../pres_elect.rds")
crime <- read_rds("../crime.rds")
votes_crime <- read_rds("../votes_crime_1979-2020_data.rds")
```


### Research question 

How does violent crime in the United Staes affect the popular vote trends across  the states?

### Cases 

**What are the cases, and how many are there?**

The cases are total violent crime data by state from 1979 through 2020, merged with presidential election vote tallies by political party for all states for the same time period.

```{r glimps_data}
glimpse(votes_crime)
```



### Data collection 

**Describe the method of data collection.**

The data for presidential elections by states was downloaded from MIT Election Data and Science Lab website. The link to the data is below:

https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/42MVDX#

The violent crime rate data by state was downloaded from Federal Bureau of Investigation Crime Data Explorer website. The link to the data is below:

https://usafacts.org/data/topics/security-safety/crime-and-justice/crime-and-police/violent-crimes/?utm_source=google&utm_medium=cpc&utm_campaign=ND-StatsData

### Type of study 

**What type of study is this (observational/experiment)?**

This is an observation study of historical data.

### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

Please see links to the data and websites utilized:

https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/42MVDX#


https://usafacts.org/data/topics/security-safety/crime-and-justice/crime-and-police/violent-crimes/?utm_source=google&utm_medium=cpc&utm_campaign=ND-StatsData


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The dependant response variable, is the party that von the popular vote in each state. This is a qualitative variable.


### Independent Variable(s)

The independant variable is the violent crime totals in each year by state. This is a quantitative variable.


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

The summary statistics are total popular votes as a percentage and crime data for the entire country. Please see plots below.

```{r crime_vote_plots}
vote_summ <- pres_elect %>%
  select(year, DEMOCRAT:REPUBLICAN) %>%
  pivot_longer(-year, names_to = "party", values_to = "votes") %>%
  group_by(year, party) %>%
  summarise(
    tot_votes = sum(votes, na.rm = TRUE),
    mean_votes = mean(votes, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(vote_pct = tot_votes / sum(tot_votes))

crime_summ <- crime %>%
  group_by(year) %>%
  summarise(
    tot_violent_crime = sum(tot_violent_crime, na.rm = TRUE),
    mean_violent_crime = mean(tot_violent_crime, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_line(
    data = crime_summ,
    aes(year, tot_violent_crime)
  )

ggplot() +
  geom_col(
    data = vote_summ,
    aes(year, vote_pct)
  ) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~party, scales = "free_y")
vote_summ
crime_summ
```
