---
title: "HW_15"
author: "Alex Khaykin"
date: "2024-05-12"
output: html_document
---

```{r}
library(cubature)
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. 
Find the equation of the regression line for the given points.  Round any final values to the nearest hundredth, if necessary.

 ( 5.6, 8.8 ), ( 6.3, 12.4 ), ( 7, 14.8 ), ( 7.7, 18.2 ), ( 8.4, 20.8 )

```{r}
# Given data points
x <- c(5.6, 6.3, 7, 7.7, 8.4)
y <- c(8.8, 12.4, 14.8, 18.2, 20.8)

# TO Calculate means
mean_x <- mean(x)
mean_y <- mean(y)

# To compute deviations
delta_x <- x - mean_x
delta_y <- y - mean_y

# TO calculate sum of products of deviations
sum_products <- sum(delta_x * delta_y)

# To calculate slope
slope <- sum_products / sum(delta_x^2)

# To determine y-intercept
intercept <- mean_y - slope * mean_x

# The regression line equation
cat("Regression line equation: y =", round(slope, 2), "x +", round(intercept, 2), "\n")
```

2. 
Find all local maxima, local minima, and saddle points for the function given below.  Write your answer(s) in the form 
( x, y, z ).  Separate multiple points with a comma.

 f ( x, y ) = 24x  6xy2  8y3

```{r}
# Define the function
f <- function(x, y) {
  return(24*x - 6*x*y^2 - 8*y^3)
}

# Calculate the partial derivatives
df_dx <- function(x, y) {
  return(24 - 6*y^2)
}

df_dy <- function(x, y) {
  return(-12*x*y - 24*y^2)
}

# Solve for critical points
y_values <- c(2, -2)
x_values <- numeric(length(y_values))

for (i in 1:length(y_values)) {
  y <- y_values[i]
  x_values[i] <- uniroot(function(x) df_dx(x, y), interval = c(-10, 10))$root
}

# Evaluate second partial derivatives
d2f_dx2 <- function(x, y) {
  return(-12*y)
}

d2f_dy2 <- function(x, y) {
  return(-12*x - 48*y)
}

# Classify critical points
for (i in 1:length(x_values)) {
  x <- x_values[i]
  y <- y_values[i]
  
  D <- d2f_dx2(x, y) * d2f_dy2(x, y)
  
  if (D > 0) {
    cat("Local maximum: (x, y) = (", x, ", ", y, ")\n")
  } else if (D < 0) {
    cat("Local minimum: (x, y) = (", x, ", ", y, ")\n")
  } else {
    cat("Saddle point: (x, y) = (", x, ", ", y, ")\n")
  }
}
```

3. 
A grocery store sells two brands of a product, the "house" brand and a "name" brand.  The manager estimates that if she 
sells the "house" brand for x dollars and the "name" brand for y dollars, she will be able to sell 81  21x + 17y units of the 
"house" brand and 40 + 11x  23y units of the "name" brand.

 Step 1. Find the revenue function R ( x, y ).
 
 Step 2. What is the revenue if she sells the "house" brand for $2.30 and the "name" brand for $4.10?

```{r}
# Define the revenue function
revenue <- function(x, y) {
  q_h <- 81 - 21 * x + 17 * y
  q_n <- 40 + 11 * x - 23 * y
  p_h <- 2.30  # Price for "house" brand
  p_n <- 4.10  # Price for "name" brand
  total_revenue <- p_h * q_h + p_n * q_n
  return(total_revenue)
}

# Calculate revenue for the given prices
price_house <- 2.30
price_name <- 4.10
revenue_at_given_prices <- revenue(price_house, price_name)

# Print the result
cat("Total revenue at prices $2.30 (house brand) and $4.10 (name brand): $", round(revenue_at_given_prices, 2), "\n")
```
4.
A company has a plant in Los Angeles and a plant in Denver.  The firm is committed to produce a total of 96 units of a 
product each week.  The total weekly cost is given by C(x, y) = 1/6 x2 + 1/6 y2 + 7x + 25y + 700,  where x is the number of units produced in Los Angeles and y is the number of units produced in Denver.  How many units should be produced in each plant to minimize the total weekly cost?


```{r}
# # Define the cost function
# cost <- function(x, y) {
#   return (1/6 * x^2 + 1/6 * y^2 + 7 * x + 25 * y + 700)
# }
# 
# # Partial derivatives
# dC_dx <- function(x, y) {
#   return (1/3 * x + 7)
# }
# 
# dC_dy <- function(x, y) {
#   return (1/3 * y + 25)
# }
# 
# # Solve for critical points
# library(stats)
# critical_points <- optim(c(0, 0), cost, method = "BFGS")$par
# x_optimal <- critical_points[1]
# y_optimal <- critical_points[2]
# 
# # Calculate the minimum cost
# min_cost <- cost(x_optimal, y_optimal)
# 
# # Print results
# cat("Optimal production quantities:\n")
# cat("x (Los Angeles):", round(x_optimal, 2), "\n")
# cat("y (Denver):", round(y_optimal, 2), "\n")
# cat("Minimum total weekly cost: $", round(min_cost, 2), "\n")
```

5.
Evaluate the double integral on the given region.
 
 
 
 e8x + 3y dA ; R: 2 £ x £ 4 and 2 £ y £ 4
 R
 Write your answer in exact form without decimals.

```{r}

# Define the integrand function
integrand <- function(xy) {
  x <- xy[1]
  y <- xy[2]
  return(exp(8*x + 3*y))
}

# Define the integration limits
lower <- c(2, 2)
upper <- c(4, 4)

# Perform the double integration
result <- adaptIntegrate(integrand, lower, upper)

# Extract the result
integral_value <- result$integral

# Print the result
cat("The exact value of the double integral is:", integral_value, "\n")
```


