---
title: "Data_605 HW_8"
author: "Alex Khaykin"
date: "2024-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#11
 A company buys 100 lightbulbs, each of which has an exponential lifetime of
 1000 hours. What is the expected time for the first of these bulbs to burn
 out?

```{r}
# Given mean lifetime (average) of a lightbulb
mean_lifetime <- 1000  # hours

# Calculate the rate parameter (λ) for exponential distribution
lambda <- 1 / mean_lifetime

# Expected lifetime of the first burnt bulb
expected_lifetime <- 1 / lambda

# Print the result
cat("Expected time for the first lightbulb to burn out:", expected_lifetime, "hours\n")
```

#14

Assume that X1 and X2 are independent random variables, each having an
 exponential density with parameter λ. Show that Z = X1 −X2 has density
 fZ(z) = (1/2)λe−λ|z| 

```{r}
# Given mean lifetime (average) of a lightbulb
mean_lifetime <- 1000  # hours

# Calculate the rate parameter (λ) for exponential distribution
lambda <- 1 / mean_lifetime

# Density function for Z
density_Z <- function(z) {
  0.5 * lambda * exp(-lambda * abs(z))
}

# Example: Compute density at z = 200
z_value <- 200
result <- density_Z(z_value)

# Print the result
cat("Density of Z at z =", z_value, "hours:", result, "\n")
```
#1

Let X be a continuous random variable with mean µ = 10 and variance
 σ2 = 100/3. Using Chebyshev’s Inequality, find an upper bound for the
 following probabilities:

 (a) P(|X −10| ≥ 2)
 (b) P(|X −10| ≥ 5)
 (c) P(|X −10| ≥ 9)
 (d) P(|X −10| ≥ 20)


```{r}
# Given parameters
mu <- 10
sigma_sq <- 100/3

# Calculate standard deviation
sigma <- sqrt(sigma_sq)

# Define the values of k for each part
k_a <- 2 / sigma
k_b <- 5 / sigma
k_c <- 9 / sigma
k_d <- 20 / sigma

# Calculate the probabilities using Chebyshev's Inequality
prob_a <- 1 / k_a^2
prob_b <- 1 / k_b^2
prob_c <- 1 / k_c^2
prob_d <- 1 / k_d^2

# Print the results
cat("Upper bounds for probabilities:\n")
cat("(a) P(|X - 10| ≥ 2):", prob_a, "\n")
cat("(b) P(|X - 10| ≥ 5):", prob_b, "\n")
cat("(c) P(|X - 10| ≥ 9):", prob_c, "\n")
cat("(d) P(|X - 10| ≥ 20):", prob_d, "\n")
```

