---
title: "Assignment 7"
author: "Alex Khaykin"
date: "2023-11-29"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(pacman)
p_load(tidyverse, ggplot2, forcats, palettes, treemapify)
```

## Data Sources
```{r}
applications <- read.csv("metals_and_applications.csv", stringsAsFactors=F)
applications <- applications %>% 
    select(-Source, -Category_Note) %>% 
    mutate(Critical.Mineral = toupper(Critical.Mineral)) %>% 
    rename(Mineral = Critical.Mineral)

comm_counts <- read.csv("metals_commodity_counts.csv", stringsAsFactors=F)
comm_counts <- comm_counts %>% 
  select(-Source) %>% 
  mutate(Country = toupper(Country))

imports <- read.csv("metals_net_imports.csv", stringsAsFactors=F)
imports <- imports %>% 
    select(-Source) %>% 
    rename(Country = Major_Import_Sources_2018_2021, Mineral = Commodity) %>% 
    mutate(Mineral = toupper(Mineral), Country = toupper(Country))
  
allies <- read.csv("united-states-allies-2023.csv", stringsAsFactors=F)
allies <- allies %>% 
  select(country, UnitedStatesAllies2023, unMember) %>% 
  rename(Country = country, isAlly = UnitedStatesAllies2023) %>% 
  mutate(Country = toupper(Country), isAlly = ifelse(is.na(isAlly), "No", "Yes"), unMember = ifelse(unMember == TRUE, "Yes", "No")) 
```

## Merging and cleaning
Imports -- we need to identify whether specific countries 
```{r}
minerals <- full_join(imports, applications, by = "Mineral")

countries <- left_join(comm_counts, allies, by = "Country") %>% 
  mutate(isAlly = ifelse(is.na(isAlly), "No", isAlly), unMember = ifelse(is.na(unMember), "No", unMember)) 
```


## Graphs

#### How many minerals are produced by each of the countries and are they an ally?
```{r}
countries %>% 
  ggplot(aes(x = fct_reorder(Country, Commodity_Count), y = Commodity_Count, fill = isAlly)) +
  geom_col() + 
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = c("goldenrod2", "brown4"), name = "U.S. Ally") +
  ylim(0,30) + 
  scale_y_continuous(expand = c(0, 0)) + 
  labs(y = "Number of Critical Minerals Imported by the U.S.", 
       x = "Country",
       title = "How many critical minerals are imported to the U.S. from \n each country?",
       subtitle = "Highest number of mineral imports come from China, Canada, and Germany.") + 
  theme(axis.text.y = element_text(size=6))
```

```{r}
countries %>% 
  group_by(isAlly) %>% 
  summarize(total = sum(Commodity_Count)) %>% 
  ggplot(aes(x = isAlly, y = total, fill = isAlly)) +
  geom_col() + 
  theme_bw() +
  scale_fill_manual(values = c("goldenrod2", "brown4"), name = "U.S. Ally") +
  labs(y = "Number of Critical Minerals Imported by the U.S.", 
       x = "Country",
       title = "How many critical minerals are imported from non-Allies?",
       subtitle = "Slightly more than 50% of all critical minerals are imported from non-allies to the U.S.") + 
  theme(axis.text.y = element_text(size=11))
```

## How does the U.S. use imported critical minerals?
```{r}
# every mineral graph
minerals %>% 
  group_by(Primary.Application.Collapsed) %>% 
  count() %>% 
  filter(!is.na(Primary.Application.Collapsed)) %>% 
  ggplot(aes(area = n, fill = n, label = Primary.Application.Collapsed)) +
  geom_treemap() +
  theme_void() +
  geom_treemap_text(colour = "white",
                    place = "centre",
                    grow = FALSE,
                    min.size = 4, 
                    reflow = TRUE) +
  scale_fill_palette_c(met_palettes$Greek, direction = -1) +
  labs(fill = "Number of Critical Minerals",
       title = "How does the U.S. use imported critical minerals?",
       subtitle = "Critical minerals are used across numerous important applications."
       )
```

## Which critical minerals are relied on entirely through import?
```{r}
minerals %>% 
  filter(!is.na(Net_Import_Reliance_pct_2022)) %>% 
  ggplot(aes(x = fct_reorder(Mineral, Net_Import_Reliance_pct_2022), y = Net_Import_Reliance_pct_2022, fill = Net_Import_Reliance_pct_2022)) +
  geom_col() +
  coord_flip() + 
  theme_bw() +
  scale_fill_palette_c(met_palettes$Greek, direction = -1) +
  scale_y_continuous(expand = c(0, 0)) + 
  labs(x = "Mineral",
       y = "% U.S.Import Reliance ",
       fill = "% Net Import Reliance, 2022",
       title = "Which critical minerals are relied entirely through imports?",
       subtitle = "Seventeen minerals have nearly 100% foreign reliance.") +
  theme(axis.text.y = element_text(size=5))

```

## Where do the top 17 critical minerals come from?
```{r}
top_minerals <- minerals %>% 
  filter(Net_Import_Reliance_pct_2022 > 95)

top_minerals_long <- top_minerals %>% 
  separate(Country, into=c("c1", "c3", "c4", "c5"), sep = ";") %>%  # max of 5
  select(-Primary.Applications) %>% 
  pivot_longer(!c(1:3,8), names_to = "key", values_to = "Country") %>% 
  group_by(Country) %>% 
  summarize(Count = n(), Proportion = n()/17) %>% 
  filter(!is.na(Country)) %>% 
  left_join(allies, by = "Country") %>% 
  mutate(isAlly = ifelse(is.na(isAlly), "No", isAlly), unMember = ifelse(is.na(unMember), "No", unMember)) 

top_minerals_long %>% 
  ggplot(aes(x = fct_reorder(Country, Proportion), y = Proportion, fill = isAlly)) +
  geom_bar(stat = "identity") + 
  coord_flip() + 
  theme_bw() +
  scale_fill_manual(values = c("goldenrod2", "brown4"), name = "U.S. Ally") +
  ylim(0,30) + 
  scale_y_continuous(expand = c(0, 0)) + 
  labs(y = "Proportion of Total U.S. Critical Mineral Imports with High Foreign Reliance", 
       x = "Country",
       title = "Where do the critical minerals with high foreign reliance \ncome from?",
       subtitle = "The majority come from China, followed by Germany.") + 
  theme(axis.text.y = element_text(size=6))
```

## Which applications would be most impacted if we were to lose one or more of the top foreign import critical minerals?
```{r}
# top 17 mineral graph
top_minerals %>% 
  group_by(Primary.Application.Collapsed) %>% 
  count() %>% 
  filter(!is.na(Primary.Application.Collapsed)) %>% 
  ggplot(aes(area = n, fill = n, label = Primary.Application.Collapsed)) +
  geom_treemap() +
  theme_void() +
  geom_treemap_text(colour = "white",
                    place = "centre",
                    grow = FALSE,
                    min.size = 4, 
                    reflow = TRUE) +
  scale_fill_palette_c(met_palettes$Greek, direction = -1) +
  labs(fill = "Number of Critical Minerals",
       title = "How does the U.S. use high-reliance imported minerals?",
       subtitle = "Industrial & chemical fields, followed by energy, battery, & fuel, would be most impacted by the \nloss of one or more of these high-reliance minerals."
       )

```

