---
title: "DATA_607_Assignment_6"
author: "Alex Khaykin"
date: "3/7/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# INTRODUCTION
The goal of this project is to learn how to create and read in tables in the JSON, HTML, and XML formats. In this project I selected three of my favorite books, one of which has two authors. 

## JSON
Calling in the library RJSON.
```{r}
library(rjson)
```

To read in the JSON table and convert to data-frame.
```{r}
my_fav_books <- fromJSON(file = "my_three_books.JSON")
my_fav_books <- as.data.frame(my_fav_books)
my_fav_books
```

## HTML
To load the package XML.
```{r}
library(XML)
```

To read in the HTML table and convert to data-frame. 
```{r}
my_fav_books2 <- readHTMLTable(doc = "my_three_books.html", as.data.frame = TRUE)
my_fav_books2
```
## XML
To read in the XML table and convert to data-frame. 
```{r}
my_fav_books3 <- xmlToDataFrame(doc = "my_three_books.xml")
my_fav_books3
```

# CONCLUSION
Comparing the JSON to HTML and XM, once converted to data-frame all are identical. Further, however please note I had to name the columns identically and in the same case for the column headers to be the same. 

