---
title: "DATA_607 Week_3 Assignmenty"
author: "Alex Khaykin"
date: "2023-02-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#INTRODUCTION
  In this assignment I will work to extract string data from a fivethirtyeight article. I will create a table and pull out majors pertaining to "DATA or "Statistic"

```{r}
library(tidyverse)
```

#1. Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"

```{r}
college_majors <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv")
```

```{r}
college_majors %>% filter(str_detect(Major, "DATA|STATISTICS")) %>%
  pull(Major) %>% c()
```

#2 Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry"

```{r}
frutas <- data.frame(c("bell pepper", "bilberry", "blackberry","blood orange","blueberry","cantalope","chili pepper","cloudberry","elderberry","lime","lychee","mulberry","olive","salal berry"))
cat(paste0(frutas))
```

#3 Describe, in words, what these expressions will match:

(.)\1\1 = to match a pattern in a strings with character repeats three times or more


"(.)(.)\\2\\1" =  match strings with a set of 4 characters, also to match with 2 characters in the same 2 characters in reverse


(..)\1 = to match any strings that have a repeated pair of letters


"(.).\\1.\\1" = to match any strings that have a repeated pair of characters


"(.)(.)(.).*\\3\\2\\1" =  to match any character strings with any 3 characters that repeat in the reverse


#4 Construct regular expressions to match words that:

Start and end with the same character:

"^(.)((.*\\1$)|\\1?$)"

```{r}
exp <- c("bob", "mom", "dad", "son", "baby", "drool", "pool", "spool", "cool", "shush", "banana")
```

```{r}
str_subset(exp, "^(.)((.*\\1$)|\\1?$)")
```

Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.):

"([A-Za-z][A-Za-z]).*\\1"

```{r}
str_subset(exp,"([A-Za-z][A-Za-z]).*\\1")
```

Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)
"([A-Za-z]).*\\1.*\\1"

```{r}
str_subset(exp, "([A-Za-z]).*\\1.*\\1")
```

