---
title: "Data 605 HW_11"
author: "Alex Khaykin"
date: "2024-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# HW Question

Using the “cars” dataset in R, build a linear model for stopping distance as a function of speed and 
replicate the analysis of your textbook chapter 3 (visualization, quality evaluation of the model, and 
residual analysis.)


```{r cars}
# Load the cars dataset
data(cars)

# summary of the data
summary(cars)
```

```{r}
# Create a scatter plot
plot(cars, xlab = "Speed (mph)", ylab = "Stopping distance (ft)", las = 1, xlim = c(0, 25))
```

```{r}
# Fit the linear model
lm_model <- lm(dist ~ speed, data = cars)

# Summary of the model
summary(lm_model)
```
```{r}
# Assess model quality (R-squared value)
cat("R-squared value:", summary(lm_model)$r.squared, "\n")

# Residual analysis
par(mfrow = c(1, 2))  # Set up a 1x2 grid for plots
plot(lm_model, which = 1)  # Residuals vs. Fitted
plot(lm_model, which = 2)  # Normal Q-Q plot
```

# Conclusion

A higher R-squared indicates a better fit.

A well-fitted model should have random residuals without any discernible structure.
