---
title: "Data HW_12"
author: "Alex Khaykin"
date: "2024-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The attached who.csv dataset contains real-world data from 2008.  The variables included follow.

Country:  name of the country 
LifeExp:  average life expectancy for the country in years 
InfantSurvival:  proportion of those surviving to one year or more 
Under5Survival:  proportion of those surviving to five years or more 
TBFree:  proportion of the population without TB. 
PropMD:  proportion of the population who are MDs 
PropRN:  proportion of the population who are RNs 
PersExp:  mean personal expenditures on healthcare in US dollars at average exchange rate 
GovtExp:  mean government expenditures per capita on healthcare, US dollars at average exchange rate 
TotExp:  sum of personal and government expenditures. 

1. Provide a scatterplot of LifeExp~TotExp, and run simple linear regression.  Do not transform the 
variables.  Provide and interpret the F statistics, R^2, standard error,and p-values only.  Discuss 
whether the assumptions of simple linear regression met. 

2. Raise life expectancy to the 4.6 power (i.e., LifeExp^4.6).  Raise total expenditures to the 0.06 
power (nearly a log transform, TotExp^.06). Plot LifeExp^4.6  as a function of TotExp^.06, and r 
re-run the simple regression model using the transformed variables.  Provide and interpret the F 
statistics, R^2, standard error, and p-values.   Which model is "better?"

3. Using the results from 3, forecast life expectancy when TotExp^.06 =1.5.  Then forecast life 
expectancy when TotExp^.06=2.5.

4. Build the following multiple regression model and interpret the F Statistics, R^2, standard error, 
and p-values.  How good is the model? 
LifeExp = b0+b1 x PropMd + b2 x TotExp  +b3 x PropMD x TotExp 

5.   Forecast LifeExp when PropMD=.03 and TotExp = 14.  Does this forecast seem realistic?  Why 
or why not? 

```{r}
# to load the dataset
who_data <- read.csv("C:\\Users\\akhay\\OneDrive\\Documents\\DATA_SCIENCE\\DATA 605\\Week_12\\who_data.csv")

# to create a scatterplot
plot(who_data$TotExp, who_data$LifeExp, xlab = "Total Expenditures", ylab = "Life Expectancy",
     main = "Scatterplot: Life Expectancy vs. Total Expenditures")

# to run simple linear regression
lm_result <- lm(LifeExp ~ TotExp, data = who_data)

# Eto extract relevant statistics
summary(lm_result)$coefficients
```


```{r}
# to transform variables
who_data$LifeExp_transformed <- who_data$LifeExp^4.6
who_data$TotExp_transformed <- who_data$TotExp^0.06

# to create a scatterplot
plot(who_data$TotExp_transformed, who_data$LifeExp_transformed,
     xlab = "Transformed Total Expenditures", ylab = "Transformed Life Expectancy",
     main = "Scatterplot: Transformed Life Expectancy vs. Transformed Total \n Expenditures")

# to run regression with new variables
lm_result_transformed <- lm(LifeExp_transformed ~ TotExp_transformed, data = who_data)

# to extract relevant statistics
summary(lm_result_transformed)$coefficients
```

```{r}
# to forecast life expectancy at TotExp_transformed = 1.5
forecast_1.5 <- predict(lm_result_transformed, newdata = data.frame(TotExp_transformed = 1.5))

# to forecast life expectancy at TotExp_transformed = 2.5
forecast_2.5 <- predict(lm_result_transformed, newdata = data.frame(TotExp_transformed = 2.5))
```


```{r}
# to build multiple regression model
lm_multiple <- lm(LifeExp ~ PropMD + TotExp + PropMD * TotExp, data = who_data)

# to extract relevant statistics
summary(lm_multiple)$coefficients
```


Smaller p-values mean more significant relationships. Also, the signs of the coefficients (+/-) indicate the direction of the relationship between the predictor and dependant variables.

```{r}
# to forecast LifeExp when PropMD = 0.03 and TotExp = 14
forecast_specific <- predict(lm_multiple, newdata = data.frame(PropMD = 0.03, TotExp = 14))
```


```{r}
summary(forecast_specific)
```

# Conclusion

The forecasted life expectancy of 108 does not seam realistic. But certainly possible.

