---
title: "Data 605 Discussion_9"
author: "Alex Khaykin"
date: "2024-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 9

#13

Physicists say that particles in a long tube are constantly moving back and
 forth along the tube, each with a velocity Vk (in cm/sec) at any given moment
 that is normally distributed, with mean µ = 0 and variance σ2 = 1. Suppose
 there are 1020 particles in the tube.
 
 (a) Find the mean and variance of the average velocity of the particles.
 (b) What is the probability that the average velocity is ≥ 10−9 cm/sec?


```{r}
# Given parameters
n_particles <- 1020
mu <- 0
sigma_sq <- 1

# Calculate mean and variance of average velocity
mu_avg <- mu / sqrt(n_particles)
sigma_avg_sq <- sigma_sq / n_particles

# Calculate Z-score for v = 10^-9 cm/sec
v <- 1e-9
Z_score <- (v - mu_avg) / sqrt(sigma_avg_sq)

# Calculate cumulative probability
prob_greater_than_v <- 1 - pnorm(Z_score)

# Display results
cat("Mean of average velocity (µ_avg):", mu_avg, "cm/sec\n")
cat("Variance of average velocity (σ_avg²):", sigma_avg_sq, "\n")
cat("Probability that average velocity ≥ 10⁻⁹ cm/sec:", prob_greater_than_v, "\n")
```

