---
title: "Data 605 HW_9"
author: "Alex Khaykin"
date: "2024-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 9

#11

The price of one share of stock in the Pilsdorff Beer Company (see Exer
cise 8.2.12) is given by Yn on the nth day of the year. Finn observes that
 the differences Xn = Yn+1 −Yn appear to be independent random variables
 with a common distribution having mean µ = 0 and variance σ2 = 1/4. If
 Y1 =100, estimate the probability that Y365 is
 
 (a) ≥100.
 (b) ≥110.
 (c) ≥120


```{r}
# Given parameters
Y_1 <- 100
variance_X <- 1/4
days <- 364

# Calculate the standard deviation of the sum of differences
sigma_S <- sqrt(days * variance_X)

# Calculate the mean and standard deviation of Y_365
mu_Y <- Y_1
sigma_Y <- sqrt(sigma_S^2)

# Calculate the standardized value for each case
Z_a <- (100 - mu_Y) / sigma_Y
Z_b <- (110 - mu_Y) / sigma_Y
Z_c <- (120 - mu_Y) / sigma_Y

# Calculate the probabilities using the cumulative distribution function (pnorm)
prob_a <- 1 - pnorm(Z_a)
prob_b <- 1 - pnorm(Z_b)
prob_c <- 1 - pnorm(Z_c)

# Output the probabilities
cat("Probability that Y_365 is greater than or equal to 100:", prob_a, "\n")
cat("Probability that Y_365 is greater than or equal to 110:", prob_b, "\n")
cat("Probability that Y_365 is greater than or equal to 120:", prob_c, "\n")
```
2.Calculate the expected value and variance of the binomial distribution using the moment 
generating function. 
3.Calculate the expected value and variance of the exponential distribution using the moment 
generating function. 

```{r}
# Given parameters
n <- 365  # Number of trials
p <- 0.5  # Probability of success

# Expected value (mean) of Y365 (binomial)
E_X <- n * p
cat("Expected value (mean) of Y365 (binomial):", E_X, "\n")

# Variance of Y365 (binomial)
Var_X <- n * p * (1 - p)
cat("Variance of Y365 (binomial):", Var_X, "\n")

# Given parameter
lambda_exp <- 1 / 365  # Rate parameter (average rate of events per day)

# Expected value (mean) of Y365 (exponential)
E_X_exp <- 1 / lambda_exp
cat("Expected value (mean) of Y365 (exponential):", E_X_exp, "\n")

# Variance of Y365 (exponential)
Var_X_exp <- 1 / lambda_exp^2
cat("Variance of Y365 (exponential):", Var_X_exp, "\n")
```


