---
title: "Data_605_HW_2"
Author: "Alex KhaYKIN"
output: html_document
date: "2024-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1A.
Proof that $A^TA \neq AA^T$

```{r}
A <- matrix(seq(from=1,to=9), nrow=3, ncol=3, byrow=T)
A_T <- t(A)
A
A_T
```
```{r}
A%*%A_T
```
```{r}
A_T%*%A
```
The matrices do not appear equal, the final test:
```{r}
A%*%A_T == A_T%*%A
```
Therefore $A^TA \neq AA^T$.

1B. Proof that $A^TA = AA^T$. T
The identity matrix is a symmetric matrix with 1s on the diagnal and zero everywhere else. 

```{r}
identity <- diag(3)
identity
```
Transpose identity matrix.

```{r}
transpose_i <- t(identity)
transpose_i
```

```{r}
identity %*% transpose_i == transpose_i %*% identity
```
Proof, that $A^TA = AA^T$
The transpose of the identity matrix or any symmetric matrix is the matrix itself.

2.

```{r}
# Set the desired number of rows and columns
N <- 5
M <- 5

# Create an N by M matrix of independent normal random variables
random_matrix <- matrix(round(runif(N * M, min = 1, max = 16), 0), nrow = N)

# View the generated matrix
print(random_matrix)
```
Writing an LU matrix factorization function. Reducing each row by an elimination amount to create an L matrix that started as an identity matrix and is used here to record the row operations performed on the U matrix.


```{r}
# LU factorization function
lu_factorization <- function(A) {
  n <- nrow(A)
  L <- diag(n)  # Initialize L as an identity matrix
  U <- A
  
  for (r in 1:n) {
      elim <- -U[r, r]/U[r,]
      
      U[r,] <- U[r,] + elim * U[r,]
      L[r,] <- elim
  }

  return(list(L = L, U = U))
}
lu_factorization(random_matrix)
```
The resulting L matrix has 1s on the diagnal and all other values record the elimination steps performed on U.
