---
title: "Data 605 HW5"
Author: Alex Khaykin
date: "2024-02-14"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

To remove and unload the EBImage package and HTML Tools
```{r}
remove.packages("htmltools")
```

Loading libraries       
```{r}
library(jpeg)
library(EBImage)
library(factoextra)
files=list.files(path='C:\\Users\\akhay\\OneDrive\\Documents\\DATA_SCIENCE\\DATA 605\\Week_4\\HW_4\\images',pattern="\\.jpg")
```

Reading in the images
```{r}
#the parameters must match the image 
height=1200
width=2500
scale=20

im=array(rep(0,length(files)*height/scale*width/scale*3),
         #set dimension to N, x, y, 3 colors, 4D array)
         dim=c(length(files), height/scale, width/scale,3)) 

#Looping and iterating over the images in the file path
for (i in 1:length(files)){
  tmp=paste0("C:\\Users\\akhay\\OneDrive\\Documents\\DATA_SCIENCE\\DATA 605\\Week_4\\HW_4\\images\\", files[i])

  temp=EBImage::resize(readJPEG(tmp),height/scale, width/scale)
  #assign to the array
  im[i,,,]=array(temp,dim=c(1, height/scale, width/scale,3))
}
```

Creating the function
```{r}
#creating the function, lots of help thank you professor
plot_shoes = function(path, add=FALSE)
{
  require('jpeg')
  jpg = readJPEG(path, native=T) # read the file
  res = dim(jpg)[2:1] # get the resolution, [x is 2, y is 1]
  if (!add) # initialize an empty plot area if add==FALSE
    plot(1,1,xlim=c(1,res[1]),ylim=c(1,res[2]), #set the X Limits by size
         asp=1, #aspect ratio
         type='n', #don't plot
         xaxs='i',yaxs='i',#prevents expanding axis windows +6% as normal
         xaxt='n',yaxt='n',xlab='',ylab='', # no axes or labels
         bty='n') # no box around graph
  rasterImage(jpg,1,1,res[1],res[2]) #image, xleft,ybottom,xright,ytop
}
```

To Plot
```{r}
par(mfrow=c(2,2)) #set graphics to 3 x 3 table
par(mai=c(.3,.3,.3,.3)) #set margins 
for (i in 1:17){  #plot the first images only
plot_shoes(writeJPEG(im[i,,,]))
}
```
To get the scalar with eigenvalues
```{r}
height=1200
width=2500
scale=20
newdata=im
dim(newdata)=c(length(files),height*width*3/scale^2)
mypca=princomp(t(as.matrix(newdata)), scores=TRUE, cor=TRUE)
sum(mypca$sdev^2/sum(mypca$sdev^2))
```
The percent variation explained by the principle components.Using a scree plot to show 80% of the variation is represented by components 1 and 2, approximately.
```{r}
mycomponents=mypca$sdev^2/sum(mypca$sdev^2) 
mycomponents
fviz_eig(mypca)
sum(mycomponents[1:3]) #Components 1 through 3 represent 85% of the variation.
```
The essence of the shoe is described nearly 70% by component 1
```{r}
mypca2=t(mypca$scores)
par(mfrow=c(3,3)) #set graphics to 3 x 3 table
par(mai=c(.3,.3,.3,.3)) #set margins 

dim(mypca2)=c(length(files),height/scale,width/scale,3)

for (i in 1:17){  #plot the first 81 Eigenshoes only
plot_shoes(writeJPEG(mypca2[i,,,], quality=1,bg="white"))
}
```

