---
title: "Data 605 HW_7"
author: "Alex Khaykin"
date: "2024-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#1 

Let X1, X2, . . . , Xn be n mutually independent random variables, each of 
which is uniformly distributed on the integers from 1 to k. Let Y denote the 
minimum of the Xi’s. Find the distribution of Y .

```{r}
# Parameters
n <- 10  # Number of random variables (change as needed)
k <- 20  # Upper bound of the uniform distribution (change as needed)

# CDF of Y
CDF_Y <- function(y) {
  if (y >= 1 && y <= k) {
    return(1 - ((k - y) / k)^n)
  } else {
    return(0)
  }
}

# PDF of Y
PDF_Y <- function(y) {
  if (y >= 1 && y <= k) {
    return(n * (k - 1) / (k - y)^(n + 1))
  } else {
    return(0)
  }
}

# Expected value of Y
E_Y <- (k + n) / (n + 1)

# Print results
cat("Distribution of Y (Beta(1, n + 1)):\n")
cat("CDF(Y <= y) =", CDF_Y(5), "\n")
cat("PDF(y) =", PDF_Y(5), "\n")
cat("E(Y) =", E_Y, "\n")
```

#2.  

Your organization owns a copier (future lawyers, etc.) or MRI (future doctors).  
This machine has a manufacturer’s expected lifetime of 10 years.  This means 
that we expect one failure every ten years.  (Include the probability statements 
and R Code for each part.). 

a.  What is the probability that the machine will fail after 8 years?.  Provide also the 
expected value and standard deviation.  Model as a geometric.  (Hint: the probability is 
equivalent to not failing during the first 8 years..) 

b.  What is the probability that the machine will fail after 8 years?.  Provide also the 
expected value and standard deviation.  Model as an exponential.   

c.  What is the probability that the machine will fail after 8 years?.  Provide also the 
expected value and standard deviation.  Model as a binomial.  (Hint:  0 success in 8 
years)   

d.  What is the probability that the machine will fail after 8 years?.  Provide also the 
expected value and standard deviation.  Model as a Poisson.  

```{r}
# Geometric Distribution
p <- 1/10
prob_geometric <- 1 - (1 - p)^8
mean_geometric <- 1 / p
sd_geometric <- sqrt((1 - p) / p^2)

# Exponential Distribution
lambda <- 1/10
prob_exponential <- exp(-lambda * 8)
mean_exponential <- 1 / lambda
sd_exponential <- 1 / lambda

# Binomial Distribution
n <- 8
prob_binomial <- (1 - p)^n
mean_binomial <- n * p
sd_binomial <- sqrt(n * p * (1 - p))

# Poisson Distribution
prob_poisson <- exp(-lambda * 8)
mean_poisson <- lambda * 8
sd_poisson <- sqrt(lambda * 8)

# Print results
cat("Geometric Distribution:\n")
cat("P(X > 8) =", prob_geometric, "\n")
cat("E(X) =", mean_geometric, "\n")
cat("SD(X) =", sd_geometric, "\n\n")

cat("Exponential Distribution:\n")
cat("P(T > 8) =", prob_exponential, "\n")
cat("E(T) =", mean_exponential, "\n")
cat("SD(T) =", sd_exponential, "\n\n")

cat("Binomial Distribution:\n")
cat("P(X = 0) =", prob_binomial, "\n")
cat("E(X) =", mean_binomial, "\n")
cat("SD(X) =", sd_binomial, "\n\n")

cat("Poisson Distribution:\n")
cat("P(X = 0) =", prob_poisson, "\n")
cat("E(X) =", mean_poisson, "\n")
cat("SD(X) =", sd_poisson, "\n")
```

