---
title: "DATA_607 PROJECT_2 DOg_Bite_Data"
author: "Alex Khaykin"
date: "3/1/2023"
output: html_document
---

# INTRODUCTION:
Dog attacks are of a particular interest to me personally, as someone who has been attacked by a vicious doberman. However, the data is very messy and untidy. I will attempt to clean it up by removing missing data, and converting all dog ages into years. The data analysis will focus on looking for a relationship between dog attacks and gender. As well as dog attacks and status of spay neuter. 

To call in knitr.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To call in the libraries.
```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
```

To read in the data.
```{r}
dog_bite <- read_csv("DOHMH_Dog_Bite_Data.csv") 
dog_bite <- as_tibble(dog_bite)
```

# Cleaning the dataset
If the the data contains a just a number, we assume it is already in years. If the data contains "M", its months, and similarly if the data contains a "W" its in weeks.

```{r}
dog_bite$Age <- gsub("Y", "", dog_bite$Age)

for(i in 1:nrow(dog_bite)){
  #print(i)
#dealing with months that were given as a range.
  if( grepl("[0-9]-[0-9]M", dog_bite[i, "Age"], ignore.case = TRUE)){
temp = gsub("-[0-9]M", "", dog_bite[i, "Age"])
temp = as.numeric(temp)
dog_bite[i, "Age"] <- as.character(round(temp/12, 3))
}
#dealing with the remaining months.
if( grepl("M", dog_bite[i, "Age"], ignore.case = TRUE)){
temp = gsub("M", "", dog_bite[i, "Age"])
temp = as.numeric(temp)
dog_bite[i, "Age"] <- as.character(round(temp/12, 3))}
#dealing with weeks.
if( grepl("W", dog_bite[i, "Age"], ignore.case = TRUE)){
temp = gsub("W", "", dog_bite[i, "Age"])
temp = as.numeric(temp)
dog_bite[i, "Age"] <- as.character(round(temp/52, 3))
}}
```

#Visualization
Is there a relationship between age, gender assigned at birth, Spayed status.

```{r}
ggplot(dog_bite, aes(y = as.numeric(Age), x = Gender, fill = Gender)) +
geom_boxplot() + ylab("Dog Age in Years")
```

```{r}
ggplot(dog_bite, aes(y = as.numeric(Age), x = SpayNeuter, fill = Gender)) +
geom_boxplot() + ylab("Dog Age in Years")
```

#CONCLUSION
In conclusion the data the there seems to be a relationship between gender and attacks. More male dogs a responsible for more attacks then female dogs, however no clear relationship exists between status of spay neuter an attacks.