---
title: "Data 605 HW_10"
author: "Alex Khaykin"
date: "2024-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Smith is in jail and has 1 dollar; he can get out on bail if he has 8 dollars. 
A guard agrees to make a series of bets with him. If Smith bets A dollars, 
he wins A dollars with probability .4 and loses A dollars with probability .6. 
Find the probability that he wins 8 dollars before losing all of his money if:

(a) he bets 1 dollar each time (timid strategy). 

(b) he bets, each time, as much as possible but not more than necessary to 
bring his fortune up to 8 dollars (bold strategy). 

(c) Which strategy gives Smith the better chance of getting out of jail? 


```{r}
# Timid Strategy
p_win <- 0.4
p_lose <- 0.6
n_bets <- 7

# Probability of winning 8 dollars
prob_timid <- p_win^n_bets
cat("Timid strategy probability:", prob_timid, "\n")

# Bold Strategy
max_dollars <- 8
prob_bold <- 0

for (i in 1:max_dollars) {
  prob_bold <- prob_bold + (p_lose^(i-1)) * p_win
}

cat("Bold strategy probability:", prob_bold, "\n")

# Comparison
if (prob_bold > prob_timid) {
  cat("Bold strategy gives Smith a better chance of getting out of jail.\n")
} else {
  cat("Timid strategy gives Smith a better chance of getting out of jail.\n")
}
```

